$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-07-27 paper/paper.tex
Add example of conversion from names to de Bruijn and back

In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: they are not inherently well-scoped, one has to deal with alpha-conversion, and inserting new variables into a context entails proving that their names differ from all other names in context.
Instead, we use de Bruijn indices \cite{deBruijn1972}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago.
That is, binders no longer introduce names; terms at different \emph{depths} use different indices to refer to the same binding.
\begin{example}
  Example process using names (top) and de Bruijn indices (bottom).
  \small
  \begin{equation*}
    \begin{aligned}
      & (\new{}x ) && (\comp {\recv{x}{y} && \send{y}{a} && \PO} {(\new{} y) && (\send{x}{y} && \recv{y}{z} && \PO)}) \\
      & \new{} && (\comp {\recv{0}{} && \send{0}{a} && \PO} {\new{} && (\send{1}{0} && \recv{0}{} && \PO)}) \\
    \end{aligned}
  \end{equation*}
\end{example}

@@

\paragraph{From names to de Bruijn indices}
We keep the original binder names around as name hints when we translate to processes with de Bruijn names --- these names will serve as hints for when we translate back.
The translation function $\func{fromRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps the first $n$ indices to their names.
Named references within the process are substituted with their corresponding de Bruijn index.
We demand that the original process is well-scoped: that all its free variable names appear in $ctx$ --- this is decidable and we therefore automate it.
\begin{equation*}
  \begin{aligned}
    &\func{fromRaw} && : (ctx : \Names_n) \; (P : \Raw) \\
    &               && \to \WellScoped \; ctx \; P \to \Process_n \\
  \end{aligned}
\end{equation*}

\paragraph{From de Bruijn indices to names}
The translation function $\func{toRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps the first $n$ indices to their names.
As many widely-used languages do, this translation function produces unique variable names.
These unique variable names use the naming scheme $<name-hint>\textasciicircum<n>$, where $<n>$ denotes that the name $<name-hint>$ has already been bound $n$ times before.
\begin{equation*}
  \begin{aligned}
    \func{toRaw} : (ctx : \Names_n) \to \Process_n \to \Raw
  \end{aligned}
\end{equation*}

>>>>

In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: they are not inherently well-scoped, one has to deal with alpha-conversion, and inserting new variables into a context entails proving that their names differ from all other names in context.
Instead, we use de Bruijn indices \cite{deBruijn1972}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago.
That is, binders no longer introduce names; terms at different \emph{depths} use different indices to refer to the same binding.

@@

\begin{example}
  Example process using names (top) translated to de Bruijn indices (middle) and back to names (bottom). The result is a process that follows the Barendregt convention and is $\alpha$-equivalent to the original.
  \small
  \begin{equation*}
    \begin{aligned}
      & (\new{x} ) && (\comp {\recv{x}{x} && \send{x}{z} && \PO} {(\new{y}) && (\send{x}{y} && \recv{y}{y} && \PO)}) \\
      & \new{} && (\comp {\recv{0}{} && \send{0}{2} && \PO} {\new{} && (\send{1}{0} && \recv{0}{} && \PO)}) \\
      & (\new{x^0} ) && (\comp {\recv{x^0}{x^1} && \send{x^1}{z^0} && \PO} {(\new{y^0}) && (\send{x^0}{y^0} && \recv{y^0}{y^1} && \PO)}) \\
    \end{aligned}
  \end{equation*}
\end{example}

\paragraph{From names to de Bruijn indices}
When we translate into de Bruijn indices we keep the original binder names around --- they will serve as name hints for when we translate back.
The translation function $\func{fromRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps the first $n$ indices to their names.
Named references within the process are substituted with their corresponding de Bruijn index.
We demand that the original process is well-scoped: that all its free variable names appear in $ctx$ --- this is decidable and we therefore automate the construction of such proof term.
\begin{equation*}
  \begin{aligned}
    &\func{fromRaw} && : (ctx : \Names_n) \; (P : \Raw) \\
    &               && \to \WellScoped \; ctx \; P \to \Process_n \\
  \end{aligned}
\end{equation*}

\paragraph{From de Bruijn indices to names}
The translation function $\func{toRaw}$ works recursively, keeping a context $ctx : \Names_n$ that maps the first $n$ indices to their names.
As some widely-used languages do, this translation function produces unique variable names.
These unique variable names use the naming scheme $<name-hint>^{<n>}$, where $^{<n>}$ denotes that the name $<name-hint>$ has already been bound $n$ times before.
\begin{equation*}
  \begin{aligned}
    \func{toRaw} : (ctx : \Names_n) \to \Process_n \to \Raw
  \end{aligned}
\end{equation*}
